Code used to generate paper results : Database comparison and CT characteristics

### 2024-05-10 Paper_code-Database_CT.Rmd
Update and enhancement on database comparison

```{r load_data, warning=FALSE, message=FALSE}
GTEX_data <- CTdata::GTEX_data()
normal_tissues_multimapping_data <- CTdata::normal_tissues_multimapping_data()
CCLE_data <- CTdata::CCLE_data()
TCGA_TPM <- CTdata::TCGA_TPM()
DAC_treated_cells_multimapping <- CTdata::DAC_treated_cells_multimapping()
testis_sce <- CTdata::testis_sce()
scRNAseq_HPA <- CTdata::scRNAseq_HPA()
CT_mean_methylation_in_tissues <- CTdata::CT_mean_methylation_in_tissues()
####
load("../CT_genes_latest_version.rda")
CT_genes$testis_specificity <- "testis_specific"
```

@@

@@

To characterise the differences between our database and other, we need the 
category we created in CTexploreR.

Hereunder is what we used for our selection pipeline (coming from 
`make_CT_list.R` in `CTdata`), mainly how we created the data.

```{r make_CT, message = FALSE, warning = FALSE}
# GTEX data with the tissue specificity category determined
all_genes <- as_tibble(rowData(GTEX_data), rownames = "ensembl_gene_id")
all_genes$TPM_testis <- assay(GTEX_data)[, "Testis"]

# Add multimapping_analysis assessing testis-specificity of genes "lowly_expressed"

all_genes <- all_genes %>%
  left_join(as_tibble(rowData(normal_tissues_multimapping_data), 
                      rownames = "ensembl_gene_id"))

# Summarise both specificity

all_genes <- all_genes %>%
  mutate(testis_specificity = case_when(
    GTEX_category == "testis_specific" |
      multimapping_analysis == "testis_specific" ~ "testis_specific",
    GTEX_category == "testis_preferential" ~ "testis_preferential"))

# Add testis scRNA seq highest expressed cell type

all_genes <- all_genes %>%
  left_join(as_tibble(rowData(testis_sce)) %>%
              dplyr::select(external_gene_name, testis_cell_type))

# Add higher in somatic scRNAseq data of normal tissues from the Human Protein Atlas

all_genes <- all_genes %>%
  left_join(as_tibble(rowData(scRNAseq_HPA), rownames = "ensembl_gene_id") %>%
              dplyr::select(ensembl_gene_id, external_gene_name,
                            Higher_in_somatic_cell_type))

# CCLE database analysis added

all_genes <- all_genes %>%
  left_join(as_tibble(rowData(CCLE_data), rownames = "ensembl_gene_id"))
all_genes[is.na(all_genes$CCLE_category), "CCLE_category"] <- "not_in_CCLE"

# TCGA database analysis added

all_genes <- all_genes %>%
  left_join(as_tibble(rowData(TCGA_TPM), rownames = "ensembl_gene_id") %>%
              dplyr::select(ensembl_gene_id, external_gene_name, percent_pos_tum,
                            percent_neg_tum, max_TPM_in_TCGA, TCGA_category))

all_genes[all_genes$lowly_expressed_in_GTEX == TRUE &
            all_genes$multimapping_analysis == "testis_specific",
          "TCGA_category"] <- "multimapping_issue"

# Add DAC analysis
induced <- as_tibble(rowData(DAC_treated_cells_multimapping), 
                     rownames = "ensembl_gene_id") %>%
  filter(induced == TRUE) %>%
  pull(external_gene_name)

all_genes <- all_genes %>%
  mutate(DAC_induced = case_when(external_gene_name %in% induced ~ TRUE,
                                 !external_gene_name %in% induced ~ FALSE))

# Add the most biologically relevant transcript (canonical and coordinates)

ensembl <- biomaRt::useDataset("hsapiens_gene_ensembl", useMart("ensembl"))
attributes_vector <- c("ensembl_gene_id",
                       "external_gene_name",
                       "ensembl_transcript_id",
                       "external_transcript_name",
                       "chromosome_name",
                       "strand",
                       "transcript_start",
                       "transcript_end",
                       "transcription_start_site",
                       "transcript_length",
                       "transcript_biotype",
                       "transcript_is_canonical")
transcripts_infos <- as_tibble(biomaRt::getBM(attributes = attributes_vector,
                                              mart = ensembl))
canonical_transcripts <- transcripts_infos %>%
  filter(transcript_is_canonical == 1) %>%
  filter(chromosome_name %in% c(1:22, "X", "Y", "MT")) %>%
  filter(transcript_biotype == "protein_coding" | transcript_biotype == "lncRNA")
all_genes <- all_genes %>%
  left_join(canonical_transcripts %>%
              dplyr::select(ensembl_gene_id,
                            external_transcript_name, ensembl_transcript_id,
                            chromosome_name, strand, transcript_start,
                            transcript_end, transcription_start_site,
                            transcript_length, transcript_biotype))

From there, we filtered based on the testis_specificity ("testis_specific" or 
"testis_preferential"), testis_cell_type (not "Macrophage", "Endothelial",
"Myoid", "Sertoli", "Leydig"), Higher_in_somatic_cell_type (TRUE), 
CCLE_category ("activated") and TCGA_category ("activated" or 
"multimapping_issue") to have our CT genes. Then the most relevant 
transcript was validated manually (and check that there is a transcript 
activated in testis and the same is activated in tumor).

@@

We find `r length(CTdatabase_ours@IntersectionSets[["11"]])/nrow(CTdatabase_cleaned)*100`
% of CTdatabase in CTexploreR, which is `r length(CTdatabase_ours@IntersectionSets[["11"]])/nrow(CT_genes)*100`
% of our database.

@@

# 12 Genes are lost because not in any database

CTdatabase_lost[is.na(CTdatabase_lost$testis_specificity), ]$testis_specificity <- "not_specific"
table(CTdatabase_lost$testis_specificity)

table(CTdatabase_lost$testis_cell_type)

table(CTdatabase_lost$Higher_in_somatic_cell_type)

@@

`r (nrow(CTdatabase_lost) - table(CTdatabase_lost$testis_specificity)[["testis_specific"]])/nrow(CTdatabase_lost)*100`% of 
these genes are not testis specific.

@@

@@

4 in all, 59 in at least 3 databases

@@

Jamin_lost[is.na(Jamin_lost$testis_specificity), ]$testis_specificity <- "not_specific"
table(Jamin_lost$testis_specificity)

table(Jamin_lost$testis_cell_type)

table(Jamin_lost$Higher_in_somatic_cell_type)

@@

table(Jamin_lost$CCLE_category)
```

@@

@@

Wang_lost[is.na(Wang_lost$testis_specificity), ]$testis_specificity <- "not_specific"
table(Wang_lost$testis_specificity)

table(Wang_lost$testis_cell_type)

table(Wang_lost$Higher_in_somatic_cell_type)

@@

@@

@@

Carter_lost[is.na(Carter_lost$testis_specificity), ]$testis_specificity <- "not_specific"
table(Carter_lost$testis_specificity)

table(Carter_lost$testis_cell_type)

table(Carter_lost$Higher_in_somatic_cell_type)

@@

table(Carter_lost$CCLE_category)
```

@@

@@

Bruggeman_lost[is.na(Bruggeman_lost$testis_specificity), ]$testis_specificity <- "not_specific"
table(Bruggeman_lost$testis_specificity)

table(Bruggeman_lost$testis_cell_type)

table(Bruggeman_lost$Higher_in_somatic_cell_type)

@@

table(Bruggeman_lost$CCLE_category)
```

@@

@@

not_specific <- filter(lost, is.na(testis_specificity))

@@

somatic_testis <- filter(lost, testis_cell_type %in% c( "Macrophage", 
                                                        "Endothelial", "Myoid",
                                                        "Sertoli", "Leydig"))

testis_expression(somatic_testis$external_gene_name, cells = "all")

somatic_tissue <- filter(lost, Higher_in_somatic_cell_type == TRUE)

HPA_cell_type_expression(somatic_tissue$external_gene_name)

@@

We have lost `r length(lost_list)` genes in total. Among them, 
`r dim(not_specific)[1]/length(lost_list)*100`% are not considered testis 
specific, `r dim(somatic_testis)[1]/length(lost_list)*100`% are expressed in 
testis somatic cells, `r dim(somatic_tissue)[1]/length(lost_list)*100`% are 
expressed in somatic cells, `r dim(not_TCGA_activated)[1]/length(lost_list)*100`% 
are not activated in TCGA samples, 
`r dim(not_CCLE_activated)[1]/length(lost_list)*100`% are not activated in CCLE 
cell lines and `r transcript_prob[1]/length(lost_list)*100`% is lost due to 
transcripts problems.

@@

table(new$testis_specificity)
table(new$X_linked, new$regulated_by_methylation)
TCGA_expression(tumor = "all", genes = new$external_gene_name, 
                units = "log_TPM")
TCGA_expression(tumor = "all", 
                genes = filter(new, X_linked & regulated_by_methylation)$external_gene_name, 
                units = "log_TPM")
```
There are `r dim(new)[1]` new CT genes in CTexploreR. These are mainly testis
specific and on autosomes. Regulation by methylation is not the majority of 
them. This makes sens as other databases have been focused on regulation by 
methylation, so as here we don't segregate with that we can find these. There 
is only 11 new "major" CT that are on the X chromosome and regulated by 
methylation. CT45 are not that new.

@@

```{r all_meth}
genes <- CT_genes$external_gene_name
database <- CT_mean_methylation_in_tissues[rownames(CT_mean_methylation_in_tissues) %in% genes]

@@

We used these data to determine in which testis cell type
each gene is mostly expressed.

```{r main_cell_type_expression_barplot}
CT_genes$main_cell_type_expression <- factor(CT_genes$testis_cell_type,
                                                levels = c("SSC", 
                                               "Spermatogonia", 
                                               "Early_spermatocyte",
                                               "Late_spermatocyte",
                                               "Round_spermatid",
                                               "Elongated_spermatid",
                                               "Sperm1",
                                               "Sperm2",
                                               "Sertoli"))

CT_genes %>% 
  mutate(regulated_by_methylation = ifelse(regulated_by_methylation,
                                           "Regulated by methylation",
                                           "Not regulated by methylation")) %>%
  filter(!is.na(testis_cell_type)) %>%
  filter(main_cell_type_expression != "Sertoli") %>% 
  ggplot(aes(x = chr, fill = main_cell_type_expression,
             color = main_cell_type_expression)) +
  scale_fill_manual(values = c("lightyellow2", "moccasin", "gold", "orange","red2",
                               "darkred", "violet", "purple")) +
  scale_color_manual(values = c("lightyellow3", "navajowhite2", "gold", "orange","red2",
                               "darkred", "violet", "purple")) +
  geom_bar(stat = 'count') +
  facet_grid(main_cell_type_expression ~ regulated_by_methylation) +
  xlab("Chromosome") +
  ylab("Number of genes") +
  theme_bw() +
  theme(
    axis.text.x = element_text(color = "black", angle = 90, vjust = 0.5, size = 10),
    axis.title = element_text(size = 10, color = "gray10"),
    strip.text.y = element_blank())

```{r escapees}
CT_genes %>% 
  filter(X_linked) %>% 
  filter(testis_cell_type %in% c("Late_spermatocyte","Round_spermatid", 
                                 "Elongated_spermatid", "Sperm1", "Sperm2"))

These genes are on the X chromosome but escape X chromosome inactivation and are
activates post-meioticly.

>>>>

```{r load_data, warning=FALSE, message=FALSE}
# CCLE_data <- CTdata::CCLE_data()
# TCGA_TPM <- CTdata::TCGA_TPM()
# testis_sce <- CTdata::testis_sce()
####

load("../CTdata/eh_data/CCLE_data.rda")
load("../CTdata/eh_data/TCGA_TPM.rda")
load("../CTdata/eh_data/testis_sce.rda")
load("../CTdata/eh_data/mean_methylation_in_tissues.rda")

load("../CTdata/eh_data/all_genes.rda")
load("../CTdata/eh_data/HPA_cell_type_specificities.rda")
load("../CTdata/eh_data/CT_genes.rda")

CT_and_CTP_genes <- CT_genes
CT_genes <- (filter(CT_genes, CT_gene_type == "CT_gene"))
```

@@

# Pipeline figure values

Here is the code used to have all the values in the pipeline figure

```{r pipeline figures value}
table(all_genes$GTEX_category)
table(all_genes$multimapping_analysis)

table(all_genes$testis_specificity)
specific_to_pref <- all_genes %>%
  filter(GTEX_category == "testis_specific" |
           multimapping_analysis == "testis_specific") %>% 
  filter(testis_specificity == "testis_preferential")
table(specific_to_pref$not_detected_in_somatic_HPA)
table(specific_to_pref$CCLE_category)
table(specific_to_pref$TCGA_category)

all_genes %>% 
  filter(testis_specificity != "not_testis_specific") %>% 
  filter(CCLE_category == "activated" & 
           (TCGA_category == "multimapping_issue" | 
              TCGA_category == "activated")) %>% 
  pull(testis_specificity) %>% 
  table()

table(all_genes$CT_gene_type) #Before and after TSS verif
```

@@

To characterise the differences between our database and other, we need the 
category we created in CTexploreR. For this, we have the object `all_genes` in 
CTdata that contains the CT analysis for all genes. More info in

Hereunder is what we used for our selection pipeline (coming from 
`make_all_genes_prelim.R` and `130_make_all_genes_and_CT_genes.R` in `CTdata`).

```{r all_genes, message = FALSE, warning = FALSE}
all_genes
```

From there, we filtered based on the testis_specificity ("testis_specific", which
is based on expression in health tissue and scRNA seq info from HPA), 
CCLE_category ("activated") and TCGA_category ("activated" or 
"multimapping_issue") to have our CT genes. Then, when wanting to validate TSS 
manually, we realised that for some genes, reads were not properly aligned to
exons which might reflect a poorly defined transcription in these regions and 
are hence likely unreliable.

Some genes were also characterized as Cancer-Testis preferential genes when 
testis specificity was less stringent

@@

We find `r length(CTdatabase_ours@IntersectionSets[["11"]])/nrow(CTdatabase_cleaned)*100`
% of CTdatabase in CTexploreR, which is 
`r length(CTdatabase_ours@IntersectionSets[["11"]])/nrow(CT_genes)*100`% of our 
database.

@@

# 9 Genes are lost because not in any database

table(CTdatabase_lost$testis_specificity)

table(CTdatabase_lost$CT_gene_type)

table(CTdatabase_lost$not_detected_in_somatic_HPA)

@@

CTdatabase_lost_upset <- 
  list(`Not testis specific` = 
         filter(CTdatabase_lost,
                testis_specificity != "testis_specific")$external_gene_name,
       `Not tumour activated` = 
         filter(CTdatabase_lost,
                (TCGA_category != "activated" &
                  TCGA_category != "multimapping_issue")|
                   CCLE_category != "activated")$external_gene_name,
       `CT preferential` =
         filter(CTdatabase_lost,
                CT_gene_type == "CTP_gene")$external_gene_name)

upset_CTdatabase <- fromList(CTdatabase_lost_upset)
upset(upset_CTdatabase)

 (nrow(CTdatabase_lost) - table(CTdatabase_lost$testis_specificity)[["testis_specific"]])/nrow(CTdatabase_lost)*100
```

`r (nrow(CTdatabase_lost) - table(CTdatabase_lost$testis_specificity)[["testis_specific"]])/nrow(CTdatabase_lost)*100`
% of these genes are not testis specific.

However `r table(CTdatabase_lost$CT_gene_type)[["CTP_gene"]]` of these lost 
genes are flagged as Cancer-Testis preferential in our analysis.

@@

In their analysis, they had characterised gene specificity, some being not 
available, not found in testis, testis-restricted, testis-selective and 
testis/brain-restricted. Let's see how the lost genes qualify as they didn't
mention those were strictly testis specific.

```{r CTdatabase_classification}
CTdatabase_cleaned %>% 
  filter(external_gene_name %in% CTdatabase_lost$external_gene_name) %>% 
  pull(Classification) %>% 
  table()
```

We can see that most of them had no info or were testis-selective (I couldn't
find on website or paper how they selected categories).

@@

4 in all, 60 in at least 3 databases

@@

table(Jamin_lost$testis_specificity)

table(Jamin_lost$CT_gene_type)

table(Jamin_lost$not_detected_in_somatic_HPA)

@@

table(Jamin_lost$CCLE_category)

table(Jamin_lost$TCGA_category, Jamin_lost$CCLE_category)

Jamin_lost_upset <- 
  list(`Not testis specific` = 
         filter(Jamin_lost,
                testis_specificity != "testis_specific")$external_gene_name,
       `Not tumour activated` = 
         filter(Jamin_lost,
                (TCGA_category != "activated" &
                  TCGA_category != "multimapping_issue")|
                   CCLE_category != "activated")$external_gene_name,
       `CT preferential` =
         filter(Jamin_lost,
                CT_gene_type == "CTP_gene")$external_gene_name)

upset_Jamin <- fromList(Jamin_lost_upset)
upset(upset_Jamin)
```

@@

However `r table(Jamin_lost$CT_gene_type)[["CTP_gene"]]` of these lost 
genes are flagged as Cancer-Testis preferential in our analysis.

@@

table(Wang_lost$testis_specificity)

table(Wang_lost$CT_gene_type)

table(Wang_lost$not_detected_in_somatic_HPA)

@@

table(Wang_lost$TCGA_category, Wang_lost$CCLE_category)

Wang_lost_upset <- 
  list(`Not testis specific` = 
         filter(Wang_lost,
                testis_specificity != "testis_specific")$external_gene_name,
       `Not tumour activated` = 
         filter(Wang_lost,
                (TCGA_category != "activated" &
                  TCGA_category != "multimapping_issue")|
                   CCLE_category != "activated")$external_gene_name,
       `CT preferential` =
         filter(Wang_lost,
                CT_gene_type == "CTP_gene")$external_gene_name)

upset_Wang <- fromList(Wang_lost_upset)
upset(upset_Wang)
```

@@

However `r table(Wang_lost$CT_gene_type)[["CTP_gene"]]` of these lost 
genes are flagged as Cancer-Testis preferential in our analysis.

@@

table(Carter_lost$testis_specificity)

table(Carter_lost$CT_gene_type)

table(Carter_lost$not_detected_in_somatic_HPA)

@@

table(Carter_lost$CCLE_category)

table(Carter_lost$TCGA_category, Carter_lost$CCLE_category)

Carter_lost_upset <- 
  list(`Not testis specific` = 
         filter(Carter_lost,
                testis_specificity != "testis_specific")$external_gene_name,
       `Not tumour activated` = 
         filter(Carter_lost,
                (TCGA_category != "activated" &
                  TCGA_category != "multimapping_issue")|
                   CCLE_category != "activated")$external_gene_name,
       `CT preferential` =
         filter(Carter_lost,
                CT_gene_type == "CTP_gene")$external_gene_name)

upset_Carter <- fromList(Carter_lost_upset)
upset(upset_Carter)
```

@@

However `r table(Carter_lost$CT_gene_type)[["CTP_gene"]]` of these lost 
genes are flagged as Cancer-Testis preferential in our analysis.

@@

table(Bruggeman_lost$testis_specificity)

table(Bruggeman_lost$CT_gene_type)

table(Bruggeman_lost$not_detected_in_somatic_HPA)

@@

table(Bruggeman_lost$CCLE_category)

table(Bruggeman_lost$TCGA_category, Bruggeman_lost$CCLE_category)

Bruggeman_lost_upset <- 
  list(`Not testis specific` = 
         filter(Bruggeman_lost,
                testis_specificity != "testis_specific")$external_gene_name,
       `Not tumour activated` = 
         filter(Bruggeman_lost,
                (TCGA_category != "activated" &
                  TCGA_category != "multimapping_issue")|
                   CCLE_category != "activated")$external_gene_name,
       `CT preferential` =
         filter(Bruggeman_lost,
                CT_gene_type == "CTP_gene")$external_gene_name)

upset_Bruggeman <- fromList(Bruggeman_lost_upset)
upset(upset_Bruggeman)
```

@@

However `r table(Bruggeman_lost$CT_gene_type)[["CTP_gene"]]` of these lost 
genes are flagged as Cancer-Testis preferential in our analysis.

@@

not_specific <- filter(lost, testis_specificity == "not_testis_specific")

@@

somatic_testis <- filter(lost, not_detected_in_somatic_HPA == FALSE)

testis_expression(somatic_testis$external_gene_name, cells = "all")
HPA_cell_type_expression(somatic_testis$external_gene_name)

@@

`r length(common)` genes in our CTexploreR database are found in at least one of
the other database, which represents `r length(common)/dim(CT_genes)[1] * 100`%.

We have lost `r length(lost_list)` genes in total. Among them, 
`r dim(not_specific)[1]/length(lost_list)*100`% are not considered testis 
specific, `r dim(somatic_testis)[1]/length(lost_list)*100`% are expressed in 
somatic cells, `r dim(not_TCGA_activated)[1]/length(lost_list)*100`% 
are not activated in TCGA samples, 
`r dim(not_CCLE_activated)[1]/length(lost_list)*100`% are not activated in CCLE 
cell lines and `r transcript_prob[1]/length(lost_list)*100`% is lost due to 
transcripts problems.

@@

table(new$testis_specificity)
table(new$X_linked)
table(new$regulated_by_methylation)
table(new$X_linked, new$regulated_by_methylation)
TCGA_expression(tumor = "all", genes = new$external_gene_name, 
                units = "log_TPM")
TCGA_expression(tumor = "all", 
                genes = filter(new, X_linked & regulated_by_methylation)$external_gene_name, 
                units = "log_TPM")
```

There are `r dim(new)[1]` new CT genes in CTexploreR. These are all testis
specific and mainly on autosomes. Regulation by methylation is the majority of 
them.  There 
is only 11 new "major" CT that are on the X chromosome and regulated by 
methylation. CT45 are not that new.

@@

```{r all_meth}
genes <- CT_genes$external_gene_name
database <- mean_methylation_in_tissues[rownames(mean_methylation_in_tissues) %in% genes]

@@
